D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-02-01 01intro_preliminary.tex
.

Roughly speaking, encoding in plain Dirac notation requires
tensoring identity operators and employing additional $\Swap$ gates,
since generally the conventional order cannot guarantee 1. the order
of all local operations is consistent with it, 2. the local operations
involve only adjacent qubits in the conventional order.
\end{example}
To address the limitations of Dirac notations, physicists routinely
use labels (or subscripts) to indicate the systems on which quantum
states or operations are applied, thereby avoiding unnecessary lifting
and swap gates. For example, rewriting the above example using labels
yields:
\begin{align*}
  \ccnot_{prq}|\ghz\rangle_{pqr} = 
  \cnot_{rq}|\ghz\rangle_{pqr}
  = \cnot_{pq}|\ghz\rangle_{pqr}
\end{align*}
This formalization avoids determining and maintaining the conventional
order of qubits, nor lifting using additional $I$ and $\Swap$s. In
this setting, the tensor products become associative and
commutative, so we can rearrange qubits as needed to
perform calculations. For our example, we can perform the calculation as
follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\rangle_{pqr}
  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
  = (|000\rangle + |110\rangle)_{prq}\\
  &\cnot_{rq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
  =(|000\rangle + |101\rangle)_{rqp}\\
  &\cnot_{pq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
  = (|000\rangle + |101\rangle)_{pqr}
\end{align*}
where the RHS of each line are equivalent.  In addition, labelled
Dirac notation can conveniently describe local measurements, partial
traces (the state or evolution of subsystems in many-body systems),
and partial inner products (the representation of partial traces in
pure states), which are sufficient to handle the mathematical formulas
of quantum mechanics in many-body systems.

Labelled Dirac notation is not only ubiquitous in the description of many-body systems, but also plays an important role in program logics. Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system. Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in [cite].

>>>>

Roughly speaking, encoding in plain Dirac notation requires
tensoring identity operators and employing additional $\Swap$ gates,
since generally the conventional order cannot guarantee 1. the order
of all local operations is consistent with it, 2. the local operations
involve only adjacent qubits in the conventional order.
\end{example}
To address the limitations of Dirac notations, physicists routinely
use labels (or subscripts) to indicate the systems on which quantum
states or operations are applied, thereby avoiding unnecessary lifting
and swap gates. For example, rewriting the above example using labels
yields:
\begin{align*}
  \ccnot_{prq}|\ghz\rangle_{pqr} = 
  \cnot_{rq}|\ghz\rangle_{pqr}
  = \cnot_{pq}|\ghz\rangle_{pqr}
\end{align*}
This formalization avoids determining and maintaining the conventional
order of qubits, nor lifting using additional $I$ and $\Swap$s. In
this setting, the tensor products become associative and
commutative, so we can rearrange qubits as needed to
perform calculations. For our example, we can perform the calculation as
follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\rangle_{pqr}
  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
  = (|000\rangle + |110\rangle)_{prq}\\
  &\cnot_{rq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
  =(|000\rangle + |101\rangle)_{rqp}\\
  &\cnot_{pq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
  = (|000\rangle + |101\rangle)_{pqr}
\end{align*}
where the RHS of each line is equivalent.  In addition, labelled
Dirac notation can conveniently describe local measurements, partial
traces (the state or evolution of subsystems in many-body systems),
and partial inner products (the representation of partial traces in
pure states), which are sufficient to handle the mathematical formulas
of quantum mechanics in many-body systems.

Labelled Dirac notation is not only ubiquitous in the description of many-body systems, but also plays an important role in program logics. Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system. Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{QSL_Le_2022}.
